---
title: "Dose-finding for Untargeted Metabolomics Studies"
author: "Yannik Schermer"
format: 
  html: 
    toc: true
bibliography: "bib.bib"
---

## Introduction

This document contains the code for the analysis of dose-finding experiments for the following compounds:

-   2-deoxy-d-glucose

-   3-bromopyruvate

-   5-fluorouracil

-   antimycin A

-   FCCP

-   menadione

-   metrizamide

-   mevastatin

-   phenanthrene-9,10-dione

-   simvastatin

-   trifluorothymidine

Data was acquired by Yannik Schermer and Frederic Wagner.

## Data Cleaning and Normalization

Raw data was cleaned and Relative viability was calculated using @eq-normalization-1

$$ Relative\:viability /\%= \frac{E_T-E_B}{E_N-E_B}\cdot100\:\%$$ {#eq-normalization-1} with

$E_T$ Extinction treated well

$E_B$ Extinction blank

$E_N$ Extinction negative control (Medium or DMSO, depending on vehicle)

## Dose-response Curve Fitting and IC~10~ Calculation

### General

A tiered approach was used to estimate IC~10~ values. First, cells were treated with a broad dosage range to determine the region of interest for a later experiment with a narrower range. From this second experiment, final IC~10~ values were calculated. Dose-response modelling is done using the `drc_pipe()` function based on the `drc` package (@ritz_dose-response_2015).

```{r,output=FALSE}
library(drc)
library(tidyverse)
library(scales)
library(egg)
```

```{r, warning=FALSE}
drc_pipe <- function(dataframe, output_list, compound_name ="", lowest_conc_CI, highest_conc_CI, lower_limit_drm = c(NA, NA, NA, NA), upper_limit_drm = c(NA, NA, NA, NA), ec_level = 10, save_plot = F, plot_name = "plot.svg", show = "all", ylim_plot = c(0, NA), y_axis_breaks = seq(0, 150, 25)){

  #model data
  drc <- dataframe %>%
    drm(viab_perc ~ c_uM, fct = LL.4(), lowerl = lower_limit_drm, upperl = upper_limit_drm, data = .)
    
  #calculate EC
  calculated_ec <- drc %>%
    ED(respLev = ec_level, interval = "delta", display = F)
  
  # For a prettier plot, data is normalized a second time using the upper and lower plateau of the first fit.
  # This ensures that upper plateau is always 100 % and the lower is 0 %.
  # Slope and Hill Coefficient are preserved to ensure that the curve shape does not change.
  # For reference, see https://www.graphpad.com/support/faq/50-of-what-how-exactly-are-ic50-and-ec50-defined/
  
  
  # Fitting normalized data.
  
  #Subtract 0 line
  upper_plateau <- unname(drc$coefficients[3])
  lower_plateau <- unname(drc$coefficients[2])
  hill_coeff <- unname(drc$coefficients[1])
  ec50 <- unname(drc$coefficients[4])
  
  dataframe$viab_perc <- dataframe$viab_perc-lower_plateau
  
  drc <- dataframe %>%
    drm(viab_perc ~ c_uM, fct = LL.4(fixed = c(hill_coeff, NA, NA, ec50)), data = .)
  
  
  #Scale to 100%
  upper_plateau <- unname(drc$coefficients[2])
  lower_plateau <- unname(drc$coefficients[1])
  
  dataframe$viab_perc <- dataframe$viab_perc/upper_plateau*100
  
  drc <- dataframe %>%
    drm(viab_perc ~ c_uM, fct = LL.4(fixed = c(hill_coeff, NA, NA, ec50)), data = .)
  
  #draw DRC with  95 %-CI
  modeled_df <- expand.grid(conc = exp(seq(log(lowest_conc_CI), log(highest_conc_CI), length = 10000)))
  modeled_df[,c(2,3,4)] <- drc %>%
    predict(newdata = modeled_df, interval = "confidence")
  
  if (show == "all") {
    show_df <- dataframe
  } else if (show == "mean"){
    show_df<- dataframe %>%
      group_by(c_uM) %>%
      summarize(viab_perc = mean(viab_perc, na.rm=TRUE))
  } else {
    print("show must be 'mean' or 'all'")
  }

  
  drc_plot <- ggplot(modeled_df, aes(conc, V2)) +
    geom_line() +
    scale_x_log10(breaks = scales::trans_breaks("log10", function(x) 10^x),
                  labels = scales::trans_format("log10", scales::math_format(10^.x)))+
    scale_y_continuous(breaks = y_axis_breaks,
                       labels = y_axis_breaks,
                       limits = ylim_plot)+
    geom_point(data = show_df, aes(c_uM, viab_perc))+
    geom_line(aes(x = conc, y = V3), alpha = 0.2, linetype = "dashed")+
    geom_line(aes(x = conc, y = V4), alpha = 0.2, linetype = "dashed")+
    theme_bw()+
    annotation_logticks(sides = "b")+
    labs(title = compound_name, x = NULL, y = NULL)
  
  output_list$EC.values[nrow(output_list$EC.values)+1,1] <- compound_name
  output_list$EC.values[nrow(output_list$EC.values),2] <- calculated_ec
  output_list$EC.values[nrow(output_list$EC.values),3] <- ec_level
  
  output_list$plots[[length(output_list$plots)+1]] <- drc_plot
  
  if (save_plot == T) {
    ggsave(filename = plot_name, plot = drc_plot, width = 75, height = 75, units = "mm", create.dir = T)
  }
  return(output_list)
}
```

Plots and calculated IC~10~ values are appended to a pre existing list by the above function. This simple helper function creates a list with the correct formatting.

```{r}
make_output_list <- function() {
  ec_values <- data.frame(
    Compound = as.character(c()),
    EC = as.numeric(c()),
    EC.level = as.numeric(c())
  )
  plots <- list()
  #dfs <- list()
  output_list <- list(ec_values, plots)
  
  names(output_list) <- c("EC.values", "plots")
  
  return(output_list)
}

output_list_range <- make_output_list()

output_list_range

```

An example for the usage of the function `drc_pipe()` is given below. Data is from menadione 2h range-finder experiment.

```{r}
menadione_range <- read.csv("files/menadione_range.csv")
knitr::kable(head(menadione_range), align = "c")

```

```{r, warning=FALSE}

output_list_range <- drc_pipe(dataframe = menadione_range, compound_name ="Menadione", lowest_conc_CI = 0.01, highest_conc_CI = 1000000, show = "mean", lower_limit_drm = c(NA, -10, 90, NA), upper_limit_drm = c(NA, 0, 110, NA), save_plot = F, plot_name = "output/plots/menadione_range.svg", output_list = output_list_range)

```

Data frames containing EC values can be accessed using the following command:

```{r, warning=FALSE}

output_list_range$EC.values

```

Dose-response can be plotted using the command below. Dose-response curve of menadione is given as an example [@fig-drc-menadione].

```{r, warning=FALSE}
#| label: fig-drc-menadione
#| fig-cap: "Dose-response curve of menadione in range finder experiments. Dashed lines represend the 95 % confidence interval."
#| warning: false

output_list_range$plots[[1]]

```

### Analysis

#### Range finder experiments

```{r, echo=FALSE}

output_list_range <- make_output_list()

```

```{r, warning=FALSE}
amca_range <- read.csv("files/amca_range.csv")

brpyr_range <- read.csv("files/brpyr_range.csv")

doglc_range <- read.csv("files/doglc_range.csv")

fccp_range <- read.csv("files/fccp_range.csv")

fu_range <- read.csv("files/fu_range.csv")

menadione_range <- read.csv("files/menadione_range.csv")

metrizamide_range <- read.csv("files/metrizamide_range.csv")

mevast_range <- read.csv("files/mevast_range.csv")

phenanthr_range <- read.csv("files/phenanthr_range.csv")

simvast_range <- read.csv("files/simvast_range.csv")

tft_range <- read.csv("files/tft_range.csv")


output_list_range <- drc_pipe(dataframe = amca_range, compound_name ="Antimycin A, Range", lowest_conc_CI = 0.01, highest_conc_CI = 1000000, show = "mean", lower_limit_drm = c(NA, -10, 90, NA), upper_limit_drm = c(NA, 0, 110, NA), save_plot = T, plot_name = "output/plots/amca_range.svg", output_list = output_list_range)


output_list_range <- drc_pipe(dataframe = brpyr_range, compound_name ="3-Bromopyruvate, Range", lowest_conc_CI = 0.01, highest_conc_CI = 1000000, show = "mean", lower_limit_drm = c(NA, -10, 90, NA), upper_limit_drm = c(NA, 0, 110, NA), save_plot = T, plot_name = "output/plots/brpyr_range.svg", output_list = output_list_range)


output_list_range <- drc_pipe(dataframe = fccp_range, compound_name ="FCCP, Range", lowest_conc_CI = 0.01, highest_conc_CI = 1000000, show = "mean", lower_limit_drm = c(NA, -10, 90, NA), upper_limit_drm = c(NA, 0, 110, NA), save_plot = T, plot_name = "output/plots/fccp_range.svg", output_list = output_list_range)


output_list_range <- drc_pipe(dataframe = menadione_range, compound_name ="Menadione, Range", lowest_conc_CI = 0.01, highest_conc_CI = 1000000, show = "mean", lower_limit_drm = c(NA, -10, 90, NA), upper_limit_drm = c(NA, 0, 110, NA), save_plot = T, plot_name = "output/plots/menadione_range.svg", output_list = output_list_range)


##mevastatin pilot 1000 µM was removed -> compound precipitated
output_list_range <- drc_pipe(dataframe = mevast_range, compound_name ="Mevastatin, Range", lowest_conc_CI = 0.01, highest_conc_CI = 1000000, show = "mean", lower_limit_drm = c(NA, -10, 90, NA), upper_limit_drm = c(NA, 0, 110, NA), save_plot = T, plot_name = "output/plots/mevastatin_range.svg", output_list = output_list_range)


output_list_range <- drc_pipe(dataframe = phenanthr_range, compound_name ="9,10-Phenanthrenedione, Range", lowest_conc_CI = 0.01, highest_conc_CI = 1000000, show = "mean", lower_limit_drm = c(NA, -10, 90, NA), upper_limit_drm = c(NA, 0, 110, NA), save_plot = T, plot_name = "output/plots/phenanthr_range.svg", output_list = output_list_range)


output_list_range <- drc_pipe(dataframe = simvast_range, compound_name ="Simvastatin, Range", lowest_conc_CI = 0.01, highest_conc_CI = 1000000, show = "mean", lower_limit_drm = c(NA, -10, 90, NA), upper_limit_drm = c(NA, 0, 110, NA), save_plot = T, plot_name = "output/plots/simvast_range.svg", output_list = output_list_range)


```

Cells treated with 2-deoxy-d-glucose, 5-fluorouracil, metrizamide and trifluorothymidine did not show reduced viability for up to 1 mM. Therefore, for metabolomics experiments, cells were treated with 1 mM of the respective compound.

```{r}
doglc_range_mean <- doglc_range %>%
  group_by(c_uM) %>%
  summarize(viab_perc = mean(viab_perc, na.rm=TRUE))

doglc_range_plot <- ggplot(doglc_range_mean, aes(c_uM, viab_perc))+
  geom_point()+
  scale_x_log10()+
  scale_y_continuous(breaks = seq(0, 150, 25),
                     labels = seq(0, 150, 25),
                     limits = c(0, NA))+
  theme_bw()+
  annotation_logticks(sides = "b")+
  labs(title = "2-Deoxy-d-glucose, Range", x = NULL, y = NULL)
ggsave(filename = "output/plots/doglc_range.svg", plot = doglc_range_plot, width = 75, height = 75, units = "mm")

output_list_range$EC.values[nrow(output_list_range$EC.values)+1,1] <- "2-Deoxy-d-glucose, Range"
output_list_range$EC.values[nrow(output_list_range$EC.values),2] <- 1000
output_list_range$EC.values[nrow(output_list_range$EC.values),3] <- NA
  
output_list_range$plots[[length(output_list_range$plots)+1]] <- doglc_range_plot




fu_range_mean <- fu_range %>%
  group_by(c_uM) %>%
  summarize(viab_perc = mean(viab_perc, na.rm=TRUE))

fu_range_plot <- ggplot(fu_range_mean, aes(c_uM, viab_perc))+
  geom_point()+
  scale_x_log10()+
  scale_y_continuous(breaks = seq(0, 150, 25),
                     labels = seq(0, 150, 25),
                     limits = c(0, NA))+
  theme_bw()+
  annotation_logticks(sides = "b")+
  labs(title = "5-Fluorouracil, Range", x = NULL, y = NULL)
ggsave(filename = "output/plots/fu_range.svg", plot = fu_range_plot, width = 75, height = 75, units = "mm")

output_list_range$EC.values[nrow(output_list_range$EC.values)+1,1] <- "5-Fluorouracil, Range"
output_list_range$EC.values[nrow(output_list_range$EC.values),2] <- 1000
output_list_range$EC.values[nrow(output_list_range$EC.values),3] <- NA
  
output_list_range$plots[[length(output_list_range$plots)+1]] <- fu_range_plot



metrizamide_range_mean <- metrizamide_range %>%
  group_by(c_uM) %>%
  summarize(viab_perc = mean(viab_perc, na.rm=TRUE))

metrizamide_range_plot <- ggplot(metrizamide_range_mean, aes(c_uM, viab_perc))+
  geom_point()+
  scale_x_log10()+
  scale_y_continuous(breaks = seq(0, 150, 25),
                     labels = seq(0, 150, 25),
                     limits = c(0, NA))+
  theme_bw()+
  annotation_logticks(sides = "b")+
  labs(title = "Metrizamide, Range", x = NULL, y = NULL)
ggsave(filename = "output/plots/metrizamide_range.svg", plot = metrizamide_range_plot, width = 75, height = 75, units = "mm")

output_list_range$EC.values[nrow(output_list_range$EC.values)+1,1] <- "Metrizamide, Range"
output_list_range$EC.values[nrow(output_list_range$EC.values),2] <- 1000
output_list_range$EC.values[nrow(output_list_range$EC.values),3] <- NA
  
output_list_range$plots[[length(output_list_range$plots)+1]] <- metrizamide_range_plot




tft_range_mean <- tft_range %>%
  group_by(c_uM) %>%
  summarize(viab_perc = mean(viab_perc, na.rm=TRUE))

tft_range_plot <- ggplot(tft_range_mean, aes(c_uM, viab_perc))+
  geom_point()+
  scale_x_log10()+
  scale_y_continuous(breaks = seq(0, 150, 25),
                     labels = seq(0, 150, 25),
                     limits = c(0, NA))+
  theme_bw()+
  annotation_logticks(sides = "b")+
  labs(title = "Trifluorothymidine, Range", x = NULL, y = NULL)
ggsave(filename = "output/plots/tft_range.svg", plot = tft_range_plot, width = 75, height = 75, units = "mm")

output_list_range$EC.values[nrow(output_list_range$EC.values)+1,1] <- "Trifluorothymidine, Range"
output_list_range$EC.values[nrow(output_list_range$EC.values),2] <- 1000
output_list_range$EC.values[nrow(output_list_range$EC.values),3] <- NA
  
output_list_range$plots[[length(output_list_range$plots)+1]] <- tft_range_plot
```

Calculated IC~10~ values (in µM):

```{r, echo=FALSE}
output_list_range$EC.values <- output_list_range$EC.values %>%
  mutate(EC = signif(EC, 3))

knitr::kable(output_list_range$EC.values, align = "c")
```

Dose-response curves of range finder experiments are shown in [@fig-drc-rangefinder-2h-all].

```{r, warning=FALSE}
#| label: fig-drc-rangefinder-2h-all
#| fig-cap: "Dose-response curves of tested compounds in range finder experiments. Dashed lines represend the 95 % confidence interval."
#| warning: false
#| fig-height: 10


ggarrange(plots = output_list_range$plots, ncol = 2, left = "Relative viability / %", bottom = "c / µM")
```

```{r, eval=FALSE}
output_list_range$dfs %>%
  bind_rows() %>%
  ggplot(aes(conc, V2))+
    geom_point()+
    facet_grid(~cpd)

```

#### Experiments to estimate the IC~10~ more precisely

After initial range finder experiments, the region of interest was estimated for each compound based on the turning point and the steepness of the curve. Compounds that did not show cytotoxicity for up to 1 mM were not tested again.

```{r, warning=FALSE}
output_list_precise <- make_output_list()

amca_precise <- read.csv("files/amca_precise.csv")

#remove possible outlier
amca_precise <- amca_precise %>%
  filter(c_uM > 0.9)

brpyr_precise <- read.csv("files/brpyr_precise.csv")

fccp_precise <- read.csv("files/fccp_precise.csv")

menadione_precise <- read.csv("files/menadione_precise.csv")

mevast_precise <- read.csv("files/mevast_precise.csv")

phenanthr_precise <- read.csv("files/phenanthr_precise.csv")

#compound precipitated <= 200 µM
phenanthr_precise <- phenanthr_precise %>%
  filter(c_uM < 200)

simvast_precise <- read.csv("files/simvast_precise.csv")
#compound precipitated <= 1000 µM
simvast_precise <- simvast_precise %>%
  filter(c_uM < 1000)







output_list_precise <- drc_pipe(dataframe = amca_precise, compound_name ="Antimycin A, Precise", lowest_conc_CI = 0.9, highest_conc_CI = 10000, show = "mean", lower_limit_drm = c(NA, -10, 90, NA), upper_limit_drm = c(NA, 0, 110, NA), save_plot = T, plot_name = "output/plots/amca_precise.svg", output_list = output_list_precise)


output_list_precise <- drc_pipe(dataframe = brpyr_precise, compound_name ="3-Bromopyrivate, Precise", lowest_conc_CI = 2, highest_conc_CI = 10000, show = "mean", lower_limit_drm = c(NA, -10, 90, NA), upper_limit_drm = c(NA, 0, 110, NA), save_plot = T, plot_name = "output/plots/brpyr_precise.svg", output_list = output_list_precise)


output_list_precise <- drc_pipe(dataframe = fccp_precise, compound_name ="FCCP, Precise", lowest_conc_CI = 2, highest_conc_CI = 10000, show = "mean", lower_limit_drm = c(NA, -10, 90, NA), upper_limit_drm = c(NA, 0, 110, NA), save_plot = T, plot_name = "output/plots/fccp_precise.svg", output_list = output_list_precise)


output_list_precise <- drc_pipe(dataframe = menadione_precise, compound_name ="Menadione, Precise", lowest_conc_CI = 9, highest_conc_CI = 10000, show = "mean", lower_limit_drm = c(NA, -10, 90, NA), upper_limit_drm = c(NA, 0, 110, NA), save_plot = T, plot_name = "output/plots/menadione_precise.svg", output_list = output_list_precise)


output_list_precise <- drc_pipe(dataframe = mevast_precise, compound_name ="Mevastatin, Precise", lowest_conc_CI = 0.9, highest_conc_CI = 10000, show = "mean", lower_limit_drm = c(NA, -10, 90, NA), upper_limit_drm = c(NA, 0, 110, NA), save_plot = T, plot_name = "output/plots/mevast_precise.svg", output_list = output_list_precise)


output_list_precise <- drc_pipe(dataframe = phenanthr_precise, compound_name ="Phenanthrene-9,10-dione, Precise", lowest_conc_CI = 1, highest_conc_CI = 10000, show = "mean", lower_limit_drm = c(NA, -10, 90, NA), upper_limit_drm = c(NA, 0, 110, NA), save_plot = T, plot_name = "output/plots/phenanthr_precise.svg", output_list = output_list_precise)


output_list_precise <- drc_pipe(dataframe = simvast_precise, compound_name ="Simvastatin, Precise", lowest_conc_CI = 2, highest_conc_CI = 10000, show = "mean", lower_limit_drm = c(NA, -10, 90, NA), upper_limit_drm = c(NA, 0, 110, NA), save_plot = T, plot_name = "output/plots/simvast_precise.svg", output_list = output_list_precise)
```

Calculated IC~10~ values (in µM):

```{r, echo=FALSE}
output_list_precise$EC.values <- output_list_precise$EC.values %>%
  mutate(EC = signif(EC, 3))

knitr::kable(output_list_precise$EC.values, align = "c")
```

```{r, warning=FALSE}
#| label: fig-drc-precise-2h-all
#| fig-cap: "Dose-response curves of more precise estimation experiments. Dashed lines represend the 95 % confidence interval."
#| warning: false
#| fig-height: 6.7
#| 
ggarrange(plots = output_list_precise$plots, ncol = 2, left = "Relative viability / %", bottom = "c / µM")
```

## Summary

Estimated IC~10~ values used for metabolomics experiments are summarized in the table below:

```{r}

```

## Session Info

```{r}
sessionInfo()
```


## References
